Imputation of Incomplete Multilevel Data with R

### 2023-12-12 docs/manuscript.qmd
Analysis case study 2, data case study 3

Mutate data to get the right data types for imputation (e.g. integer for clustering variable).
```{r}
#| prompt: true
dat <- dat |>
  mutate(name = as.integer(name))
```

@@

which should be adjusted to the appropriate `2l` methods.
```{r}
#| prompt: true
meth[c("pupil", "ct")] <- "2l.pmm"
meth[c("hypots", "tsah", "edh")] <- "2l.pmm"
```

@@

and the analysis-model outcome should be used as a predictor in all imputation models
```{r}
#| prompt: true
pred[, "mort"] <- 2
```

@@

Impute the incomplete data with
```{r}
#| prompt: true
# imp <- mice(
#   dat,
#   method = meth,
#   predictorMatrix = pred,
#   printFlag = FALSE
# )
```

Evaluate the convergence of the algorithm
```{r}
#| prompt: true
# plot_trace(imp)
```

Analyze the imputed data:
```{r}
#| prompt: true
# fit <- imp %>% 
#   with(glmer(
#     mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 | name),
#     family = "binomial"
#     )) 
```
The estimated effects after imputation are presented in Table XYZ.

@@

We start by loading the data:
```{r}
#data("data_heckman", package = "micemd")
#dat <- data_heckman
```

@@

```{r, dpi = 300,out.width="70%", fig.cap="Missing pattern", eval=TRUE}
# myplots <- lapply(1:5, function(i) {
#   ggmice::plot_pattern(setDT(Obesity)[Cluster==i])+
#   ggplot2::ggtitle(paste0("Cluster", i))
#  })
# ggarrange(myplots[[1]], myplots[[3]], nrow=1,common.legend = TRUE, legend="bottom")
```

>>>>

Mutate data to get the right data types for imputation (e.g. integer for clustering variable).
```{r}
#| prompt: true
# dat <- mutate(
#   dat,
#   name = as.integer(name))
```

This is necessary because otherwise PMM cannot be used for these factor variables.
```{r}
#| prompt: true
dat <- mutate(
  dat,
  across(everything(), as.numeric))
```

@@

which should be adjusted to the appropriate `2l` methods.
```{r}
#| prompt: true
# meth[c("pupil", "ct")] <- "2l.pmm"
# meth[c("hypots", "tsah", "edh")] <- "2l.pmm"
meth[meth == "pmm"] <- "2l.pmm"
```

@@

and the analysis-model outcome should be used as a predictor in all imputation models
```{r}
#| prompt: true
pred[, "mort"] <- 2
# pred[pred == 1] <- 2
```

@@

Impute the incomplete data with
```{r}
#| prompt: true
imp <- mice(
  dat,
  method = meth,
  predictorMatrix = pred,
  printFlag = FALSE
)
```

Evaluate the convergence of the algorithm
```{r}
#| prompt: true
plot_trace(imp)
```

Evaluate the imputed values.
```{r}
#| prompt: true
ggmice(imp, aes(.imp, pupil)) +
  geom_jitter()
```

Convert the data back to factors.
```{r}
long <- complete(imp, "long", include = TRUE)
long <- mutate(long, across(c("motor_score", "pupil", "ct"), as.factor))
imp <- as.mids(long)
```

Analyze the imputed data:
```{r}
#| prompt: true
#| warning: false
fit <- imp %>%
  with(glmer(
    mort ~ type + age + motor_score + pupil + ct + (1 | name),
    family = "binomial"
    ))
```

The estimated effects after imputation are presented in Table XYZ.
```{R}
#| prompt: true
est <- testEstimates(as.mitml.result(fit), extra.pars = TRUE)
```

@@

We start by loading the data:
```{r}
#| prompt: true
data(Obesity, package = "micemd")
```

@@

```{r}
#| prompt: true
plot_pattern(Obesity)
```

```{r}
#| prompt: true
Obesity |> 
  split(~Cluster) |>
  lapply(plot_pattern)
```

### 2023-12-12 docs/manuscript.tex
Analysis case study 2, data case study 3

\begin{tcolorbox}[enhanced jigsaw, opacityback=0, arc=.35mm, colback=white, breakable, toprule=.15mm, bottomrule=.15mm, rightrule=.15mm, leftrule=.75mm, left=2mm]

@@

\begin{verbatim}
R> dat <- dat |>
+  mutate(name = as.integer(name))
\end{verbatim}

@@

\begin{verbatim}
       name        type         age motor_score       pupil          ct 
         ""          ""          ""          ""   "polyreg"   "polyreg" 
     hypots        tsah         edh        mort 
      "pmm"       "pmm"       "pmm"          "" 
\end{verbatim}

@@

\begin{verbatim}
R> meth[c("pupil", "ct")] <- "2l.pmm"
R> meth[c("hypots", "tsah", "edh")] <- "2l.pmm"
\end{verbatim}

@@

\begin{verbatim}
R> pred[, "mort"] <- 2
\end{verbatim}

@@

{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-37-1.pdf}

@@

\begin{verbatim}
R> # imp <- mice(
R> #   dat,
R> #   method = meth,
R> #   predictorMatrix = pred,
R> #   printFlag = FALSE
R> # )
\end{verbatim}

@@

\begin{verbatim}
R> # plot_trace(imp)
\end{verbatim}

@@

\begin{verbatim}
R> # fit <- imp %>% 
R> #   with(glmer(
R> #     mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 | name),
R> #     family = "binomial"
R> #     )) 
\end{verbatim}

@@

@@

\begin{verbatim}
#data("data_heckman", package = "micemd")
#dat <- data_heckman
\end{verbatim}

@@

\begin{verbatim}
# myplots <- lapply(1:5, function(i) {
#   ggmice::plot_pattern(setDT(Obesity)[Cluster==i])+
#   ggplot2::ggtitle(paste0("Cluster", i))
#  })
# ggarrange(myplots[[1]], myplots[[3]], nrow=1,common.legend = TRUE, legend="bottom")
\end{verbatim}

>>>>

\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, breakable, toprule=.15mm, rightrule=.15mm, arc=.35mm, left=2mm, leftrule=.75mm, colback=white, opacityback=0]

@@

\begin{verbatim}
R> # dat <- mutate(
R> #   dat,
R> #   name = as.integer(name))
\end{verbatim}

This is necessary because otherwise PMM cannot be used for these factor
variables.

\begin{verbatim}
R> dat <- mutate(
+  dat,
+  across(everything(), as.numeric))
\end{verbatim}

@@

\begin{verbatim}
       name        type         age motor_score       pupil          ct 
         ""          ""          ""          ""       "pmm"       "pmm" 
     hypots        tsah         edh        mort 
      "pmm"       "pmm"       "pmm"          "" 
\end{verbatim}

@@

\begin{verbatim}
R> # meth[c("pupil", "ct")] <- "2l.pmm"
R> # meth[c("hypots", "tsah", "edh")] <- "2l.pmm"
R> meth[meth == "pmm"] <- "2l.pmm"
\end{verbatim}

@@

\begin{verbatim}
R> pred[, "mort"] <- 2
R> # pred[pred == 1] <- 2
\end{verbatim}

@@

{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-38-1.pdf}

@@

\begin{verbatim}
R> imp <- mice(
+  dat,
+  method = meth,
+  predictorMatrix = pred,
+  printFlag = FALSE
+)
\end{verbatim}

@@

{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-40-1.pdf}

Evaluate the imputed values.

\begin{verbatim}
R> ggmice(imp, aes(.imp, pupil)) +
+  geom_jitter()
\end{verbatim}

{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-41-1.pdf}

Convert the data back to factors.

\begin{verbatim}
long <- complete(imp, "long", include = TRUE)
long <- mutate(long, across(c("motor_score", "pupil", "ct"), as.factor))
imp <- as.mids(long)
\end{verbatim}

@@

\begin{verbatim}
R> fit <- imp %>%
+  with(glmer(
+    mort ~ type + age + motor_score + pupil + ct + (1 | name),
+    family = "binomial"
+    ))
\end{verbatim}

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1605}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1111}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1235}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0988}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1358}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.2222}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0741}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0741}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Estimate
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Std.Error
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
t.value
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
df
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
P(\textgreater\textbar t\textbar)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
RIV
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
FMI
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
(Intercept) & -1.892 & 0.333 & -5.682 & 11445.455 & 0.000 & 0.019 &
0.019 \\
type & -0.346 & 0.178 & -1.948 & 686356.768 & 0.051 & 0.002 & 0.002 \\
age & 0.032 & 0.002 & 19.564 & 2418.910 & 0.000 & 0.042 & 0.041 \\
motor\_score2 & -0.575 & 0.070 & -8.226 & 71295.018 & 0.000 & 0.008 &
0.008 \\
motor\_score3 & -0.891 & 0.072 & -12.440 & 66712.189 & 0.000 & 0.008 &
0.008 \\
motor\_score4 & -1.276 & 0.074 & -17.305 & 18426.087 & 0.000 & 0.015 &
0.015 \\
pupil2 & 1.315 & 0.068 & 19.339 & 110.826 & 0.000 & 0.235 & 0.204 \\
pupil3 & 0.657 & 0.075 & 8.745 & 423.463 & 0.000 & 0.108 & 0.101 \\
ct2 & 0.747 & 0.084 & 8.884 & 41.781 & 0.000 & 0.448 & 0.340 \\
ct3 & 0.766 & 0.090 & 8.535 & 12.103 & 0.000 & 1.352 & 0.631 \\
\end{longtable}

\begin{longtable}[]{@{}lr@{}}
\toprule\noalign{}
& Estimate \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Intercept\textasciitilde\textasciitilde Intercept\textbar name &
0.082 \\
\end{longtable}

@@

\begin{verbatim}
R> data(Obesity, package = "micemd")
\end{verbatim}

@@

\begin{verbatim}
R> plot_pattern(Obesity)
\end{verbatim}

{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-48-1.pdf}

\begin{verbatim}
R> Obesity |> 
+  split(~Cluster) |>
+  lapply(plot_pattern)
\end{verbatim}

\begin{verbatim}
$`1`
\end{verbatim}

{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-49-1.pdf}

\begin{verbatim}

$`2`
\end{verbatim}

{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-49-2.pdf}

\begin{verbatim}

$`3`
\end{verbatim}

{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-49-3.pdf}

\begin{verbatim}

$`4`
\end{verbatim}

{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-49-4.pdf}

\begin{verbatim}

$`5`
\end{verbatim}

{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-49-5.pdf}
